Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-07-22 manuscript.qmd

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative.](figures/fig-model-risks.svg){#fig-model-risks}

@Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III only ever crosses a 5% threshold if the upstream level II is involved too, resulting in a risk of 5% to 6%<!-- TODO: make dynamic -->. As long as the contralateral neck appears clinically node negative, however, even an advanced T-category and midline extension do not yield a risk higher than 2%.

@@

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk does not exceed 5%

@@

- LNL II when the tumor extends over the mid-sagittal line
- LNL III when LNL II is clinically, the tumor is of advanced T-category, crosses the midline, and the ipsilateral neck shows involvement of at least LNL II
- LNL IV only for severe ipsilateral involvement, advanced T-category with obvious midline extension and contralateral upstream involvement

>>>>

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig-model-risks.svg){#fig-model-risks}

@Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%. For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007]. In the figure legends, this is noted e.g. for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only crosses a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->. When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.

@@

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk stays just below 3%<!-- TODO: make dynamic -->. We only observe a risk of 5-6% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA. This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation. This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.

@@

- LNL II when the tumor extends over the mid-sagittal line
- LNL III when LNL II is clinically involved, may need confirmation via FNA
- LNL IV only when upstream level III pathologically confirmed
